Distinguishing Spin-Aligned and Isotropic Black Hole Populations With Gravitational Waves

### 2017-06-05 paper/response.txt

Thanks very much for the referee reports and editorial comments on our
submitted manuscript.  We are re-submitting, having addressed all the
comments provided.  There have been a very large number of changes to
the text and references; below we list the changes made to address
each point individually.  If you would like, we could also provide a
diff against the original submitted LaTeX that would collect all the
changes together (without the logical ordering by point).

Editor's Comments
=================

Before addressing the comments in-line below, we should note that
GW170104 has been announced between receiving referee reports and this
re-submission.  We have, at the request of the editor, incorporated it
into our calculations.  This changes the odds ratio between the
isotropic and aligned models from 1.7 sigma to 2.4 sigma, and also the
expected minimum "distinguishing power" between models with ten
additional detections from 2.9 sigma to 2.4 sigma (not a
typo---increased confidence in the isotropic models means that more
detections from the "aligned" distributions are required to invert the
odds ratio in favour of these latter models).  Very little other text
has been changed as a result of the new detection (we have added
"GW170104" in appropriate places, changed "three" to "four," etc); the
central conclusions of the paper, and the caveats to the conclusions
about low spins / kick misalignments remain basically unchanged.

Item-by-item comments follow below:

> We estimate the current length of your paper to be ~1800 words,
  which exceeds our usual limit of 1500 words for Letters with four
  display items. Keep in mind that important technical details that
  are not central to the main message of the paper can be moved into
  the Methods section (right now, titled Appendix -- it is far below
  our general guideline of 3000 words).

Through addressing the other comments provided and moving material
into the Methods section (see below), we have significantly shortened
the main text.

> Letters begin with a fully referenced paragraph, ideally of about
  200 words, aimed at readers in other disciplines. This paragraph
  starts with a 2- to 3-sentence, basic introduction to the field that
  conveys to readers the ‘state of the art’ immediately before this
  work. It continues with a 1-sentence statement of the main findings
  starting 'Here we report’ or an equivalent phrase; and finally,
  concludes with 2 to 3 sentences putting the main findings into
  general context so it is clear how the results described in the
  paper have moved the field forward. Please do not introduce or use
  *any* abbreviations/acronyms in the first paragraph, and severely
  restrict their use elsewhere, as they make the paper difficult for
  non-specialists to read. The spin parameter should certainly be
  defined when first used, but it is probably not necessary to use it
  in the first para.

We have removed the abbreviations and the spin parameter frome the
abstract, and shortened it to be below 200 words.

> Further introductory material in the main text of the paper should
  not be necessary. Please note that Letters do not have
  sections/subsections/section headings.

We have removed any section headings from the letter portion of the
text (our Methods section retains section headings).

> Each display equation uses 30-40 word equivalents, and our style is
  to avoid them in the main text *unless* they are newly derived (not
  relevant here) or crucial to the overall logical flow of the
  text. The definition (eq 1) of effective spin probably could be
  shifted to the Methods, with a brief description in words in the
  main text. I am open to persuasion that eq 2 needs to be in the main
  text, but probably it too can be shifted to the Methods.

Both equations have now been shifted to the appendix.

> The "Discussion" section is rather long-winded and inconclusive. We
  recommend shifting that to the Methods, with a brief summary
  paragraph in the main text.

We have shortened and re-arranged the discussion section
significantly.  However, we would prefer not to move what remains to
Methods; we feel that the "caveats," particularly surrounding spin
magnitudes and alignment in "field" binaries, are scientifically
important.  For example, even one of our referees over-interpreted our
results, claiming that we had shown that dynamical formation scenarios
are strongly preferred over "field" formation---but in fact, it is
certainly possible that field formation of BBH mergers produces small
spins, in which case we cannot distinguish the two cases.  Moving the
remaining discussion paragraphs (there are three, and one is quite
short) to the Methods section would de-emphasise these caveats further
and, we feel, could lead to more confusion.

> Any discussion at the end of the paper should also be brief, and not
  repeat what is already written in the initial summary
  paragraph. (The final paragraph should be omitted -- Nature papers
  do not end with summaries.)

We removed the "summary" pargraph.

> At the end of the main text document (after the main figure
  legends), there should be a section entitled "Methods", which
  provides a more detailed discussion of the additional methodological
  information that would allow other researchers to replicate the
  results (we define “Methods” quite broadly). The Methods section
  will not appear in the print version but will be fully copy-edited
  and appear online in the full-text HTML and PDF versions. The
  Methods section should be written as concisely as possible but
  should contain all elements necessary to allow interpretation and
  reproduction of the results. If there are additional references in
  the Methods section, their numbering should continue from the last
  reference in the main paper, and the list should follow the Methods
  section. If the methods require chemical structures, figures or
  tables, these should be supplied as Extended Data (see below). For
  mathematically complex methods, or methods that require an unusually
  large number of figures or tables (beyond what can be accommodated
  as Extended Data), the entire Methods section should instead be
  supplied as a separate Supplementary Information.

Our previous "Appendix" has been incorporated in a "Methods" section.
We have moved material from the main text to this section (as
indicated above and below), and have also added sections on code
availability, the accumulation of evidence with events (see below re:
GW170104), and precision of chi_eff measurements (including Extended
Data Figure 5).

> Letters can have up to 30 references in the main text; additional
  references can be cited in (and listed after) the Methods section,
  as detailed above. Each reference should have its own number, and
  you should please include final page numbers, as well as initial
  ones, unless the first ‘page’ is an article number (i.e. A&A
  beginning in 2010, ApJ beginning in 2011). Our abbreviations for the
  common astronomy journals are: Astrophys. J., Astron. J.,
  Astron. Astrophys., Mon. Not. R. Astron. Soc.. Authors' last names
  should come first, followed by their initials. Any reference to work
  in preparation should be made in the main text, and not numbered in
  the reference list. As we are now printing the titles of all
  references cited, we would appreciate it if you could include them
  in the revised version of your text. The year of publication should
  be in parentheses at the end of the line.

We have re-formatted the bibliography in line with Nature
expectations.  We now have exactly 30 references in the main text,
having eliminated some citations that were superfluous and moved
others that were technical in nature to the Methods section.  The
references eliminated (in ADS notation) and the justification are

* 1993Natur.364..421K (Kulkarni, Hut & McMillan 1993): redundant with
  Siggurdsson & Hernquist (1993).

* 1973NInfo..27...70T (Tutukov & Yungelson 1973): redundant with
  Tutukov & Yungelson 1993.

* 1997MNRAS.288..245L (Lipunov, Postnov & Prokhorov 1997): redundant
  with Tutukov & Yungelson (1993) and Belczynski, et al. (2016).

* 2016PhRvL.116x1102A (Abbott, et al. 2016, properties of merger
  GW150914): redundant with Abbott, et al. (2016), 2016PhRvX...6d1015A
  (Black holes during O1 observing run).

* 2014PhRvD..89h4006P,2014PhRvD..89f1502T,2014PhRvL.113o1101H: These
  waveform technical references are not relevant to our argument or
  discussion.

* 2016ApJS..227...14A (Abbott, et al. 2016, GW150914 rates paper
  supplemental information): gives technical information not relevant
  to our argument, and therefore redundant with 2016ApJ...833L...1A
  (GW150914 rates paper).

* 2003MNRAS.341..385P (Podsiadlowski, Rappaport & Han 2003): redundant
  with 2015ApJ...800...17F (Fragos & McClintock 2015).

> We are now requiring authors to make a statement in the Methods
  section regarding code availability, under the subheading "Code
  availability". Authors can specify that "The code used to generate x
  is available at y", or say that "We have opted not to make the code
  available because ...".  This does not apply to standard data
  reduction software (which should be named), but it does apply to
  anything that you have modelled.

We now make such a statement in Methods Section 1.

> Your Figures 5-7 should be converted to Extended Data Figure 1, ED
  Figure 2 and ED Figure 3. They should be referred to in that way in
  the text.

Done.

Referee 1
=========

> The analysis presented in the manuscript is timely, well explained
  and convincing (but see a few minor points below). The authors use
  (estimates of) the measurement uncertainties published by LIGO for
  the three events to determine the probability of observing the
  measured parallel spin values from different underlying
  populations. Multiple spin magnitude distributions are considered,
  which is useful given our very limited information on the true
  distribution. In fact the authors are likely conservative with
  allowing negative spins with respect to the orbital axis for the
  aligned case.

[This statement is part of the "summary pargraphs" at the beginning of
the report.]  In fact, the referee is confused on this point; because
we *do not* allow negative spins, we obtain such distinguishing
power.  We have tried to increase the prevalence of the text pointing
this out, in light of the referee's confusion; the relevant paragraph
is found on page 10 of the main body.

> My chief concern with the manuscript is related to the novelty of
  the results. The analysis of the existing 3 (2 + 1 likely)
  detections is indeed performed here for the first time, and it is an
  interesting result. However, the calculation regarding the number of
  events needed for a 3-5 sigma constraint is similar to two previous
  works, both referenced in the manuscript. Stevenson, Berry and
  Mandel (1703.06873) carried out a very similar analysis to the one
  in the manuscript. They established that about 5 observations would
  be sufficient to distinguish between the two extreme cases of fully
  spin-aligned or fully isotropic spin distributions, the identical
  case considered in the present manuscript. While Stevenson et
  al. use simulations as opposed to relying on the reconstructed spin
  distributions by LIGO, and they consider a constant spin (0.7) as
  opposed to the distributions adopted in the present manuscript, the
  methods and conclusions are essentially the same.

It is true that some of our conclusions are the same as Stevenson et
al., but here (as the referee points out) we have for the first time
applied such an analysis to actual detections; also, we have explored
how the distinguishability between aligned and isotropic distributions
behaves when the spin = 0.7 (optimistic) assumption of Stevenson et al
is weakened.  Finally, we explore the limits of this method when spins
are intrinsically small (< 0.2), and discuss this possibility in the
context of field binary formation theory.  We feel this work, unlike
Stevenson et al, gives the modelling community a clear target: for
field fromation scenarios to work going forward, either kicks must be
extreme enough to render the distribution approximately isotropic *or*
spins must be typically < 0.2 to escape the constraints we have placed
from the existing observations.

> Regarding the above publications [Stevenson, et al. & Vitale et
  al.], while they are cited and briefly discussed in the manuscript,
  their their presentation is somewhat lacking. Mainly, the result of
  Vitale et al. is said to carry out a similar analysis, but concludes
  that several hundred sources would be needed (see lines 166-169). In
  reality, Vitale et al. focuses on measuring the precise
  contributions of different channels, e.g. 70% vs. 20%, as opposed to
  ruling out the extreme cases in which all binaries originate from
  the same channel (as is considered in the present paper). The
  several hundred measurements will be needed for such precision
  measurements, not for ruling out extreme cases.

We have re-organised the discussion of Vitale, et al. (now Ref. 17).
Rather than occupying its own paragraph, the different conclusions
from Vitale et al are addressed where appropriate in the text.  On
page 10, where we highlight that our "resolving power" comes from the
constraint on our aligned models that chi_eff > 0, we now note that
this constraint was not imposed on Vitale's "aligned" models, and
highlight that in our work we find much higher confidence in isotropy
after 10 detections as a result (>> 5 sigma versus ~3 sigma).  And
then, on page 17 (Methods Section), where we discuss our mixture model
and the possibility that there may be multiple channels in operation,
we note that Vitale+ found that it would take O(100) detections to pin
down the mixing fraction.

> One additional concern with the analysis is that it considers
  LVT151012 in the sample of observations along with the two 5sigma
  detections. I note that it is mostly standard in the literature to
  treat this candidate as a real event, on paper it has 10% chance of
  being not astrophysical. Taking this 10% false alarm into account
  may significantly reduce the 1.7sigma limit presented in the
  manuscript.

Taking the 10% probability of being "terrestrial" in origin into
account to a first approximation reduces the accumulated evidence from
LVT151012 by about 10%, so the odds ratio changes by a similar amount;
this is not a significant change.  Incorportaing a "background"
mixture model into our analysis would significantly complicate it,
without changing the final result appreciably.

> Another minor concern is that the template based searches used by
  LIGO effectively use priors for the underlying spin distributions,
  and for the searches finding the three events only spin-aligned
  templates were used. It would be useful for the reader to briefly
  comment on whether this can alter the observed spin distribution
  from the underlying distribution.

We now comment on this possibility in Methods Section 7.

Referee 2
==========

> The paper as written is too technical and jargon laden for Nature.
  The topic and result are, in my judgement, of interest to Nature
  readers, and I recommend consideration for publication with a
  significant rewrite of the abstract, introduction and conclusion to
  make it more acceptable to a general audience.

As you will have seen, we have re-written much of the introduction and
conclusion to make it less technical and/or jargon laden.

> The effective spin, χeff, appears in the abstract without context.
  This is not appropriate for a Nature paper, whose abstract is
  targeted for the general reader.

Removed.

> The abstract suggests binary black hole spins extend to high values
  - this statement is strictly limited to stellar mass black holes,
  and is derived for single black holes accreting from a main sequence
  star or giant star.

We still note in the abstract that our conclusions are dependent on
the BBH spins extending to high values, but more completely discuss
this caveat on page 10 and 11 (final pargraph of page 10, extending
onto 11).

> Section 1: the discussion of the effective spin needs to be expanded
  and made clearer for the general reader.  There is notational
  confusion with both "a" and "χ" used to refer to spin

We have moved the equation and discussion referred to here to Methods
Section 2, and now highlight the distinctions between "a" (spin
magnitude), "chi" (spin projection on z-axis), and "chi_eff"
(mass-weighted average of chis).

> It is really necessary, for a Nature paper, to have a very brief
  discussion of the spin parametrization and that there is a true
  maximum spin for a black hole.

We give this discussion and the associated limits now in Methods
Section 2.

> Approximating the posterior distribution as gaussian is strictly
  non-physical. I believe that a (truncated) gaussian is an adequate
  approximation to the distribution, but this requires some
  justification in text.

We have noted that our distributions are actually truncated in the
main text; since no distribution extends near +/- 1, truncation has no
effect on our results.

> The motivation for the choice of P(a) needs to be expanded a bit,
  the toy models used are not physically motivated and that needs to
  be made clear. I'd even suggest a figure (for the general reader) of
  P(a) for the three distributions, if there is room. Some mention of
  the more general "low" distribution in Appendix C needs to be made
  in the main text.

We state in the text (this was also present in the last submitted
version of the paper): "These magnitude distributions are not meant to
represent any particular physical model, but rather to capture our
uncertainty about the spin magnitude distribution; neither
observations nor population synthesis codes can at this point
authoritatively suggest any particular spin distribution.  Our models,
however, allow us to see how sensitive the chi_eff distribution is to
spin alignment given uncertainties about the spin magnitudes."  We
have added a figure (Extended Data Figure 1) and some discussion
(Methods Section 2, bottom of page 13) on the models.  We now mention
"Appendix C" (now Methods Section 5) in the text on page 10, when
discussing how our model selection would degrade if spin magnitudes
are small.

> I infer that the same P(a) is used for both members of the binary?
  For binaries formed in the field, the mass accretion history for
  each binary member is somewhat different in a systematic way and any
  spin due to accretion, as opposed to natal spin, is likely to be
  systematically different, to some extent, for each component. This
  may be inconsequential, but may be prudent to mention.

We now clearly state that p(a) is the same for both members in the
binary and that this may not be true in the field due to
systematically different evolutionary histories of the binary
members.  Page 4, final pargraph.

> There needs to a mention in the main text of the mass ratio
  discussion in Appendix D.

We now mention this in the final paragraph on Page 4, when discussing
the equal-mass assumption we use in the main text.  We have also added
a section dealing with the dependence of chi_eff uncertainty on mass
ratio (Methods Section 8), which is discussed Footnote 2.

> The spin-kick discussion is inadequate: there are discussion of the
  issue in Brandt et al. 1995, Sipior et al 2002, Fryer 2003, Willems
  et al 2005, as well as a number of more recent related papers.  In
  particular, IF there are kicks, they correlate with the probability
  of the binary remaining bound and its time to coalescence, and hence
  the post-kick spin alignment is correlated with probability of
  merger, possibly in a mass-dependent way. Arguably this is beyond
  the scope of the paper, but it ought to be mentioned.

We have re-written the alignment/kick discussion for field binaries to
be more clear and concise.  It now appears in the middle paragraph on
page 11.  We feel, however, that a detailed discussion of the effect
of large kicks on rates is outside the scope of this paper (if we talk
about these details, what about the effect of large kicks in dynamical
formation scenarios---they reduce the rates further here---and what
about all the other uncertainties in binary evolution and their effect
on rates/spin correlations, etc).

> The argument in the conclusion that this paper agrees with Stevenson
  et al 2017, as opposed to Vitale et al, is somewhat misleading, as
  there is substantial overlap in authorship and methodology between
  Stevenson et al and this paper. The consistency of these results
  should not be implied to constitute independent verification of
  them.

We have eliminated this paragraph, and do not make claims about
independent verification in this current draft.

### 2017-06-05 paper/AlignedIsotropicGW.tex
As re-submitted.

\begin{abstract}
  The first direct detections of gravitational
  waves\cite{2016PhRvL.116f1102A,2016PhRvL.116x1103A,O1-BBH,PhysRevLett.118.221101}
  from merging binary black holes open a unique window into the binary
  black hole formation environment.  One promising environmental
  signature is the angular distribution of the black hole spins;
  systems formed through dynamical interactions among already-compact
  objects are expected to have isotropic spin
  orientations\cite{SigurdssonHernquist:1993,PZMcMillan:2000,Rodriguez:2015,Stone:2016,2016ApJ...832L...2R}
  whereas binaries formed from pairs of stars born together are more
  likely to have spins preferentially aligned with the binary orbital
  angular
  momentum\cite{TutukovYungelson:1993,2016Natur.534..512B,Stevenson:2017,MandeldeMink:2016,Marchant:2016}.
  We consider existing gravitational wave measurements of the binary
  effective spin, the best-measured combination of spin
  parameters\cite{O1-BBH,PhysRevLett.118.221101}, in the four likely
  binary black hole detections GW150914, LVT151012, GW151226, and
  GW170104.  If binary black hole spin magnitudes extend to high
  values we show that the data exhibit a $\OOneSigmaIsoAligned\sigma$
  ($\OOneOddsIsoAligned$ odds ratio\footnote{An odds ratio of $r$ with
    $r \ll 1$ is equivalent to $x \sigma$ with
    $x = \Phi^{-1}\left( 1 - r/2 \right)$, where $\Phi$ is the unit
    normal CDF.})  preference for an isotropic angular distribution
  over an aligned one.  By considering the effect of 10 additional
  detections\cite{2016ApJ...833L...1A}, we show that such an augmented
  data set would enable in most cases a prefernce stronger than
  $5\sigma$ ($2.9 \times 10^{-7}$ odds ratio).  The existing
  preference for either an isotropic spin distribution or low spin
  magnitudes for the observed systems will be confirmed (or
  overturned) confidently in the near future.
\end{abstract}

@@

Each of our models for the merging binary black hole spin population
assumes that the merging black holes are of equal mass (this is
marginally consistent with the
observations\cite{O1-BBH,PhysRevLett.118.221101}, and the $\chieff$
distribution is not particularly sensitive to the mass ratio between
the merging objects---see Methods Section \ref{sec:mass-ratio}).  We
assume that the population spin distribution factorises into a
distribution for the spin magnitude $a$ and a distribution for the
spin angles.  Finally, we assume that the distribution of spins is
common to each component in a merging binary (in reality the
distributions of spin for each component in the binary may differ
systematically due to different formation histories).  Choosing one of
three magnitude distributions (see Methods Section
\ref{methsec:chieff-spin-magnitude}), ``low'' (mean $a = 0.33$,
standard deviation $0.24$), ``flat'' (mean $a = 0.5$, standard
deviation $0.29$), ``high'' (mean $a = 0.67$, standard deviation
$0.24$) and pairing with either an isotropic angular distribution or a
distribution that generates perfect alignment with the positive $z$
axis yields six different models for the $\chieff$ distribution.
These models are shown in Figure \ref{fig:chieff-distribution-models}.

@@

These magnitude distributions are not meant to represent any
particular physical model, but rather to capture our uncertainty about
the spin magnitude distribution, discussed in detail below; neither
observations nor population synthesis codes can at this point
authoritatively suggest \emph{any} particular spin
distribution\cite{2015PhR...548....1M}.  Our models, however, allow us
to see how sensitive the $\chieff$ distribution is to spin alignment
given uncertainties about the spin magnitudes.

We fit hierarchical models of the three existing LIGO O1 and GW170104
observations using these six different, zero-parameter population
distributions (see Methods Section \ref{sec:hierarchical}).  We also
fit three mixture models for the population, where the spin magnitude
distribution is fixed but the angular distribution is a weighted sum
of the isotropic and aligned distribution.  The evidence, or marginal
likelihood, for each of the models is shown in Figure
\ref{fig:O1-odds}.  For all three magnitude distributions, the mixture
models' posterior on the mixing fraction peaks at 100\% isotropic,
which explains why the zero-parameter, pure-isotropic models are
preferred over the single-parameter mixture models for every magnitude
distribution with this data set.  Not surprisingly, given the small
$\chieff$ values in the three detected systems, the most-favoured
model among those with an isotropic angular distribution has the
``low'' magnitude distribution; the most favoured model among those
with an aligned distribution also has the ``low'' magnitude
distribution.  The odds ratio between the ``low'' aligned and ``low''
isotropic models is $\OOneOddsIsoAligned$, or
$\OOneSigmaIsoAligned\sigma$; thus the data favour isotropic spins
among our suite of models.  While the data favour spin amplitude
distributions with small spin magnitudes, note that a model with all
binary black hole systems having zero spin is ruled out by the
GW151226 measurements, which bound at least one black hole to have
spin magnitude $\geq 0.2$ at 99\%
credibility\cite{2016PhRvL.116x1103A}.

@@

Estimates of the rate of binary black hole coalescences give a
reasonable chance of 10 additional binary black hole detections in the
next three years\cite{O1-BBH,2016ApJ...833L...1A}.  Assuming 10
additional detections drawn from each of our six zero-parameter models
for the spin distribution in addition to the four existing detections
from O1 and GW170104, with observational uncertainties drawn randomly
from the three Gaussian widths used to approximate the $\chieff$
posteriors in Figure \ref{fig:O1-posteriors}\footnote{The measurement
  uncertainty in $\chieff$ depends on the other parameters of the
  merging binary black hole system, particularly on the mass ratio.
  Our assumption about future observational uncertainties is
  appropriate if the parameters of the three detected events are
  representative of the parameters of future detections. See Methods
  Section \ref{sec:chi-eff-precision} for further discussion.}, we
find the odds ratios shown in Figure \ref{fig:O2-predictions}.  We
find that most scenarios with an additional 10 detections allow the
simulated angular distribution to be inferred with greater than
$5\sigma$ ($2.9 \times 10^{-7}$ odds) credibility; and in the most
pessimistic case the distinction is typically
$\OTwoSigmaIsoAlignedMin\sigma$ ($\OTwoOddsIsoAlignedMin$ odds ratio).
While such future detections should permit a confident distinction
between \emph{angular} distributions, we would remain much less
certain about the \emph{magnitude} distribution among the three
options considered here until we have a larger number of observations.
In Figure \ref{fig:O2-predictions}, the odds ratio between different
magnitude distributions with the same angular distribution is much
closer to unity than the odds ratio between angular distributions.

@@

Most of our resolving power for the spin angular distribution is a
result of the fact that our ``aligned'' models cannot produce
$\chieff < 0$ (see Figure \ref{fig:chieff-distribution-models}).  If
spins are intrinsically very small, with $a \lesssim 0.2$, then it is
no longer possible to resolve the negative effective spin with a small
number of observations (see Methods Section \ref{sec:smallspins}).  As
noted below, however, spins observed in X-ray binaries are typically
large.  Additionally, models which do not permit \emph{some} spins
with $\chieff \gtrsim 0.1$ are ruled out by the GW151226
observations\cite{2016PhRvL.116x1103A}.  An ``aligned'' model with
spin magnitudes from our ``flat'' distribution but permitting spin
vectors oriented anti-parallel to the orbital angular momentum
(leading to the possibility of positive \emph{or} negative $\chieff$)
can only be distinguised from an isotropic true population at
$\sim 3 \sigma$ after 10--20 observations\cite{2017CQGra..34cLT01V};
our flat aligned model can be distinguished from such a population at
better than $5\sigma$ (odds $< 10^{-8}$) after 10 observations,
emphasising the information content of the bound $\chieff > 0$ for our
aligned models.

>>>>

\begin{abstract}
  The first direct detections of gravitational
  waves\cite{2016PhRvL.116f1102A,2016PhRvL.116x1103A,O1-BBH,PhysRevLett.118.221101}
  from merging binary black holes open a unique window into the binary
  black hole formation environment.  One promising environmental
  signature is the angular distribution of the black hole spins;
  systems formed through dynamical interactions among already-compact
  objects are expected to have isotropic spin
  orientations\cite{SigurdssonHernquist:1993,PZMcMillan:2000,Rodriguez:2015,Stone:2016,2016ApJ...832L...2R}
  whereas binaries formed from pairs of stars born together are more
  likely to have spins preferentially aligned with the binary orbital
  angular
  momentum\cite{TutukovYungelson:1993,2016Natur.534..512B,Stevenson:2017,MandeldeMink:2016,Marchant:2016}.
  We consider existing gravitational wave measurements of the binary
  effective spin, the best-measured combination of spin
  parameters\cite{O1-BBH,PhysRevLett.118.221101}, in the four likely
  binary black hole detections GW150914, LVT151012, GW151226, and
  GW170104.  If binary black hole spin magnitudes extend to high
  values we show that the data exhibit a $\OOneSigmaIsoAligned\sigma$
  ($\OOneOddsIsoAligned$ odds ratio\footnote{An odds ratio of $r$ with
    $r \ll 1$ is equivalent to $x \sigma$ with
    $x = \Phi^{-1}\left( 1 - r/2 \right)$, where $\Phi$ is the unit
    normal CDF.})  preference for an isotropic angular distribution
  over an aligned one.  By considering the effect of 10 additional
  detections\cite{2016ApJ...833L...1A}, we show that such an augmented
  data set would enable in most cases a preference stronger than
  $5\sigma$ ($2.9 \times 10^{-7}$ odds ratio).  The existing
  preference for either an isotropic spin distribution or low spin
  magnitudes for the observed systems will be confirmed (or
  overturned) confidently in the near future.
\end{abstract}

@@

Each of our models for the merging binary black hole spin population
assumes that the merging black holes are of equal mass (this is
marginally consistent with the
observations\cite{O1-BBH,PhysRevLett.118.221101}, and the $\chieff$
distribution is not sensitive to the mass ratio---see Methods Section
\ref{sec:mass-ratio}).  We assume that the population spin
distribution factorises into a distribution for the spin magnitude $a$
and a distribution for the spin angles.  Finally, we assume that the
distribution of spins is common to each component in a merging binary
(the distributions of spin for each component in the binary could
differ systematically due to different formation histories).  Choosing
one of three magnitude distributions (see Methods Section
\ref{methsec:chieff-spin-magnitude}), ``low'' (mean $a = 0.33$,
standard deviation $0.24$), ``flat'' (mean $a = 0.5$, standard
deviation $0.29$), ``high'' (mean $a = 0.67$, standard deviation
$0.24$) and pairing with an isotropic angular distribution or a
distribution that generates perfect alignment yields six different
models for the $\chieff$ distribution.  These models are shown in
Figure \ref{fig:chieff-distribution-models}.

@@

These magnitude distributions are not meant to represent any
particular physical model, but rather to capture our uncertainty about
the spin magnitude distribution; neither observations nor population
synthesis codes can at this point authoritatively suggest \emph{any}
particular spin distribution\cite{2015PhR...548....1M}.  Our models,
however, allow us to see how sensitive the $\chieff$ distribution is
to spin alignment given uncertainties about the spin magnitudes.

We fit hierarchical models of the three existing LIGO O1 and GW170104
observations using these six different, zero-parameter population
distributions (see Methods Section \ref{sec:hierarchical}).  We also
fit three mixture models for the population, where the angular
distribution is a weighted sum of the isotropic and aligned
distributions.  The evidence, or marginal likelihood, for each of the
models is shown in Figure \ref{fig:O1-odds}.  For all three magnitude
distributions, the mixture models' posterior on the mixing fraction
peaks at 100\% isotropic.  Not surprisingly, given the small $\chieff$
values in the three detected systems, the most-favoured model among
those with an isotropic angular distribution has the ``low'' magnitude
distribution; the most favoured model among those with an aligned
distribution also has the ``low'' magnitude distribution.  The odds
ratio between the ``low'' aligned and ``low'' isotropic models is
$\OOneOddsIsoAligned$, or $\OOneSigmaIsoAligned\sigma$; thus the data
favour isotropic spins among our suite of models.  While the data
favour spin amplitude distributions with small spin magnitudes, note
that a model with all binary black hole systems having zero spin is
ruled out by the GW151226 measurements, which bound at least one black
hole to have spin magnitude $\geq 0.2$ at 99\%
credibility\cite{2016PhRvL.116x1103A}.

@@

Estimates of the rate of binary black hole coalescences give a
reasonable chance of 10 additional binary black hole detections in the
next three years\cite{O1-BBH,2016ApJ...833L...1A}.  Assuming 10
additional detections drawn from each of our six zero-parameter models
in addition to the four existing detections from O1 and GW170104, with
observational uncertainties drawn randomly from the three Gaussian
widths used to approximate the $\chieff$ posteriors in Figure
\ref{fig:O1-posteriors}\footnote{The measurement uncertainty in
  $\chieff$ depends on the other parameters of the merging binary
  black hole system, particularly on the mass ratio.  Our assumption
  about future observational uncertainties is appropriate if the
  parameters of the three detected events are representative of the
  parameters of future detections. See Methods Section
  \ref{sec:chi-eff-precision} for further discussion.}, we find the
odds ratios shown in Figure \ref{fig:O2-predictions}.  We find that
most scenarios with an additional 10 detections allow the simulated
angular distribution to be inferred with greater than $5\sigma$
($2.9 \times 10^{-7}$ odds) credibility. In the most pessimistic case
the distinction is typically $\OTwoSigmaIsoAlignedMin\sigma$
($\OTwoOddsIsoAlignedMin$ odds ratio).  While such future detections
should permit a confident distinction between \emph{angular}
distributions, we would remain much less certain about the
\emph{magnitude} distribution among the three options considered here
until we have a larger number of observations.

@@

Most of our resolving power for the spin angular distribution is a
result of the fact that our ``aligned'' models cannot produce
$\chieff < 0$ (see Figure \ref{fig:chieff-distribution-models}).  If
spins are intrinsically very small, with $a \lesssim 0.2$, then it is
no longer possible to resolve the negative effective spin with a small
number of observations (see Methods Section \ref{sec:smallspins}).  As
noted below, however, spins observed in X-ray binaries are typically
large.  Additionally, models which do not permit \emph{some} spins
with $\chieff \gtrsim 0.1$ are ruled out by the GW151226
observations\cite{2016PhRvL.116x1103A}.  An ``aligned'' model with
spin magnitudes from our ``flat'' distribution but permitting spin
vectors oriented anti-parallel to the orbital angular momentum
(leading to the possibility of positive \emph{or} negative $\chieff$)
can only be distinguished from an isotropic true population at
$\sim 3 \sigma$ after 10--20 observations\cite{2017CQGra..34cLT01V};
our flat aligned model can be distinguished from such a population at
better than $5\sigma$ (odds $< 10^{-8}$) after 10 observations,
emphasizing the information content of the bound $\chieff > 0$ for our
aligned models.
